---
title: "Income Analysis(Adult Dataset)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Importing Libraries, dataset and attribute name.
```{r }
library(ggplot2)
library(plyr)
library(ROCR)
df <- read.table('adult.data', sep = ',', fill = F, strip.white = T)
colnames(df) <- c('age', 'workclass', 'fnlwgt', 'educatoin','educatoin_num', 'marital_status', 'occupation', 'relationship','race', 'sex', 'capital_gain', 'capital_loss', 'hours_per_week', 'native_country', 'income')
```
# Overview of dataset
```{r }
head(df)
```
# Structure of the dataset
```{r }
str(df)
```
# dimension of dataset
```{r }
dim(df)
```
# Removing unncessary attribute from the dataset
```{r }
df$educatoin <- NULL
df$fnlwgt <- NULL
df$relationship <- NULL
```

# histogram of age by income group
```{r }
ggplot(df) + aes(x=as.numeric(age), group=income, fill=income) + 
  geom_histogram(binwidth=1, color='black')
```

# histogram of age by gender group
```{r }
ggplot(adult) + aes(x=as.numeric(age), group=sex, fill=sex) + 
  geom_histogram(binwidth=1, color='black')
```
#The variable workclass stands for the industry in which the responding unit is employed.
```{r }
summary(df$workclass)

```
#Combining Different Groups
```{r }
levels(df$workclass)[1] <- 'Unknown'
# combine into Government job
df$workclass <- gsub('^Federal-gov', 'Government', df$workclass)
df$workclass <- gsub('^Local-gov', 'Government', df$workclass)
df$workclass <- gsub('^State-gov', 'Government', df$workclass) 

# combine into Sele-Employed job
df$workclass <- gsub('^Self-emp-inc', 'Self-Employed', df$workclass)
df$workclass <- gsub('^Self-emp-not-inc', 'Self-Employed', adult$workclass)

# combine into Other/Unknown
df$workclass <- gsub('^Never-worked', 'Other', df$workclass)
df$workclass <- gsub('^Without-pay', 'Other', df$workclass)
df$workclass <- gsub('^Other', 'Other/Unknown', df$workclass)
df$workclass <- gsub('^Unknown', 'Other/Unknown', df$workclass)

df$workclass <- as.factor(df$workclass)
```
# barplot of job type by income group
# get the counts by industry and income group
```{r }
count <- table(df[adult$workclass == 'Government',]$income)["<=50K"]
count <- c(count, table(df[df$workclass == 'Government',]$income)[">50K"])
count <- c(count, table(df[df$workclass == 'Other/Unknown',]$income)["<=50K"])
count <- c(count, table(df[df$workclass == 'Other/Unknown',]$income)[">50K"])
count <- c(count, table(df[df$workclass == 'Private',]$income)["<=50K"])
count <- c(count, table(df[df$workclass == 'Private',]$income)[">50K"])
count <- c(count, table(df[df$workclass == 'Self-Employed',]$income)["<=50K"])
count <- c(count, table(df[df$workclass == 'Self-Employed',]$income)[">50K"])
count <- as.numeric(count)

# create a dataframe
industry <- rep(levels(df$workclass), each = 2)
income <- rep(c('<=50K', '>50K'), 4)
df1 <- data.frame(industry, income, count)
df1
```
# calculate the percentages
```{r }
df1 <- ddply(df1, .(industry), transform, percent = count/sum(count) * 100)
```
# format the labels and calculate their positions
```{r }
df1 <- ddply(df1, .(industry), transform, pos = (cumsum(count) - 0.5 * count))
df1$label <- paste0(sprintf("%.0f", df$percent), "%")
```
# bar plot of counts by industry with in group proportions 
```{r }
ggplot(df1, aes(x = industry, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income by Industry')
```
# create a dataframe
```{r }
df2 <- data.frame(table(df$income, df$educatoin_num))
names(df2) <- c('income', 'education_num', 'count')
df2
```
```{r }
# calculate the percentages
df2 <- ddply(df2, .(education_num), transform, percent = count/sum(count) * 100)

# format the labels and calculate their positions
df2 <- ddply(df2, .(education_num), transform, pos = (cumsum(count) - 0.5 * count))
df2$label <- paste0(sprintf("%.0f", df2$percent), "%")

# remove some in group percentage to avoid overlapped text
df2$label[which(df2$percent < 5)] <- NA

# bar plot of counts by years of education with in group proportions 
ggplot(df2, aes(x = education_num, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level with Years of Education')
```

```{r }
summary(df$occupation)

```
```{r }
levels(df$occupation)[1] <- 'Unknown'
df$occupation <- gsub('Adm-clerical', 'White-Collar', df$occupation)
df$occupation <- gsub('Craft-repair', 'Blue-Collar', df$occupation)
df$occupation <- gsub('Exec-managerial', 'White-Collar', df$occupation)
df$occupation <- gsub('Farming-fishing', 'Blue-Collar', df$occupation)
df$occupation <- gsub('Handlers-cleaners', 'Blue-Collar', df$occupation)
df$occupation <- gsub('Machine-op-inspct', 'Blue-Collar', df$occupation)
df$occupation <- gsub('Other-service', 'Service', df$occupation)
df$occupation <- gsub('Priv-house-serv', 'Service', df$occupation)
df$occupation <- gsub('Prof-specialty', 'Professional', df$occupation)
df$occupation <- gsub('Protective-serv', 'Service', df$occupation)
df$occupation <- gsub('Tech-support', 'Service', df$occupation)
df$occupation <- gsub('Transport-moving', 'Blue-Collar', df$occupation)
df$occupation <- gsub('Unknown', 'Other/Unknown', df$occupation)
df$occupation <- gsub('Armed-Forces', 'Other/Unknown', df$occupation)
df$occupation <- as.factor(df$occupation)
summary(df$occupation)
```

# create a dataframe
```{r }
df3 <- data.frame(table(df$income, df$occupation))
names(df3) <- c('income', 'occupation', 'count')
df3
```
# calculate the percentages and their visualization 
```{r }

df3 <- ddply(df3, .(occupation), transform, percent = count/sum(count) * 100)

# format the labels and calculate their positions
df3 <- ddply(df3, .(occupation), transform, pos = (cumsum(count) - 0.5 * count))
df3$label <- paste0(sprintf("%.0f", df3$percent), "%")

# bar plot of counts by occupation with in group proportions 
ggplot(df3, aes(x = occupation, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level with Different Occupations')
```

```{r }
summary(df$marital_status)
```
#marital_status is a categorical variable with 7 categories indicating the marital status of observations. In fact, it can be blocked into a few categories as well
```{r }
df$marital_status <- gsub('Married-AF-spouse', 'Married', df$marital_status)
df$marital_status <- gsub('Married-civ-spouse', 'Married', df$marital_status)
df$marital_status <- gsub('Married-spouse-absent', 'Married', df$marital_status)
df$marital_status <- gsub('Never-married', 'Single', df$marital_status)
df$marital_status <- as.factor(df$marital_status)
summary(df$marital_status)
```
#creating a datafrom for income and martial status
```{r }
df4 <- data.frame(table(df$income, df$marital_status))
names(df4) <- c('income', 'marital_status', 'count')
head(df4)
```
# Visualization representation of Martial Status and transformation
```{r }
# calculate the percentages
df4 <- ddply(df4, .(marital_status), transform, percent = count/sum(count) * 100)

# format the labels and calculate their positions
df4 <- ddply(df4, .(marital_status), transform, pos = (cumsum(count) - 0.5 * count))
df4$label <- paste0(sprintf("%.0f", df4$percent), "%")

# bar plot of counts by marital status with in group proportions 
ggplot(df4, aes(x = marital_status, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level with Marital Status')

```
# histogram of capital_gain
```{r }

ggplot(adult) + aes(x=as.numeric(capital_gain), group=income, fill=income) + 
  geom_histogram(bins=10, color='black') + ggtitle('Histogram of Capital Gain')
```
# histogram of capital_loss
```{r }
ggplot(adult) + aes(x=as.numeric(capital_loss), group=income, fill=income) + 
  geom_histogram(bins=10, color='black') + ggtitle('Histogram of Capital Loss')
```
# percentage of observatiosn with no capital gain or loss

```{r }
sum(df$capital_gain == 0)/length(df$capital_gain)
```

#Sum of capital Los
```{r }
sum(df$capital_loss == 0)/length(df$capital_loss)
```
# Removing Capital_gain, Capital_loss, and Native_country
```{r }
df$capital_gain <- NULL
df$capital_loss <- NULL
df$native_country <- NULL
```
# data frame for race and their visualization
```{r }
df5 <- data.frame(table(df$income, df$race))
names(df5) <- c('income', 'race', 'count')
head(df5)
```

# Removing missing values from race variable, conversion of Race variable and visualization
```{r }
# calculate the percentages
df5 <- ddply(df5, .(race), transform, percent = count/sum(count) * 100)

# format the labels and calculate their positions
df5 <- ddply(df5, .(race), transform, pos = (cumsum(count) - 0.5 * count))
df5$label <- paste0(sprintf("%.0f", df4$percent), "%")

# do not display percentage for low counts categories
df5$label[df5$race == 'Other'] <- NA
df5$label[df5$race == 'Amer-Indian-Eskimo'] <- NA

# bar plot of counts by marital status with in group proportions 
ggplot(df5, aes(x = race, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level by Race')
```
# Training and Test Split Dataset

```{r }
sz <- round(.8 * dim(df)[1])  # training set size
training_set <- df[1:sz,]
testing_set <- df[-(1:sz),]
```
#Logistic Regression Model
```{r }
m1 <- glm(income ~ ., data = training_set, family = binomial('logit'))
summary(m1)
```
# confint of logistic Regression Model
```{r }
confint(m1)

```
#Feature Selection backward and forward Method
```{r }
m_full <- m1  # full model is the model just fitted
m_null <- glm(income ~ 1, data = training_set, family = binomial('logit'))

# backward selection
step(m_full, trace = F, scope = list(lower=formula(m_null), upper=formula(m_full)),
     direction = 'backward')
```
# forward Feature selection Method
```{r }
step(m_null, trace = F, scope = list(lower=formula(m_null), upper=formula(m_full)),
     direction = 'forward')
```
# create a data frame to store information regarding deviance residuals
```{r }
index <- 1:dim(training_set)[1]
dev_resid <- residuals(m1)
income <- training_set$income
dff <- data.frame(index, dev_resid, income)

ggplot(dff, aes(x = index, y = dev_resid, color = income)) +
  geom_point() + 
  geom_hline(yintercept = 3, linetype = 'dashed', color = 'blue') +
  geom_hline(yintercept = -3, linetype = 'dashed', color = 'blue')

```
# Performance Evaluation for Logistic Regression Model
```{r }
 prob <- predict(m1, testing_set, type = 'response')
pred <- rep('<=50K', length(prob))
pred[prob>=.5] <- '>50K'
# confusion matrix 
Confusion_Matrix <- table(pred, testing_set$income)
Confusion_Matrix


```
# Random Forest Model
```{r }
library(randomForest)
rf3 <- randomForest(income ~ ., data = training_set, ntree = 1000)
rf3.pred.prob <- predict(rf3, newdata = testing_set, type = 'prob')
rf3.pred <- predict(rf3, newdata = testing_set, type = 'class')
# confusion matrix 
Confusion_Matrix_RF <- table(rf3.pred, testing_set$income)
Confusion_Matrix_RF
```
# Decision Tree Model
```{r }
library(rpart)
dt <- rpart(income ~ ., data = training_set, method = 'class', cp = 1e-3)
dt.pred.prob <- predict(dt, newdata = testing_set, type = 'prob')
dt.pred <- predict(dt, newdata = testing_set, type = 'class')
# confusion matrix 
Confusion_Matrix_dt <- table(dt.pred, testing_set$income)
Confusion_Matrix_dt
```
# Receiver Operator Characteristics perfomance evaluation
```{r }
# create a prediction object
pr <- prediction(prob, testing_set$income)
prf <- performance(pr, measure = "tpr", x.measure = "fpr")

# create a data frame for TP and FP rates
dd <- data.frame(FP = prf@x.values[[1]], TP = prf@y.values[[1]])


# CART
pr2 <- prediction(dt.pred.prob[,2], testing_set$income)
prf2 <- performance(pr2, measure = "tpr", x.measure = "fpr")
dd2 <- data.frame(FP = prf2@x.values[[1]], TP = prf2@y.values[[1]])

# RF
pr3 <- prediction(rf3.pred.prob[,2], testing_set$income)
prf3 <- performance(pr3, measure = "tpr", x.measure = "fpr")
dd3 <- data.frame(FP = prf3@x.values[[1]], TP = prf3@y.values[[1]])



# plot ROC curve for logistic regression
g <- ggplot() + 
  geom_line(data = dd, aes(x = FP, y = TP, color = 'Logistic Regression')) + 
  geom_line(data = dd2, aes(x = FP, y = TP, color = 'CART')) + 
  geom_line(data = dd3, aes(x = FP, y = TP, color = 'Random Forest')) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1)) +
  ggtitle('ROC Curve') + 
  labs(x = 'False Positive Rate', y = 'True Positive Rate') 
g +  scale_colour_manual(name = 'Classifier', values = c('Logistic Regression'='#E69F00','Random Forest'='#D55E00', 'CART'='#009E73'))

```
# Area under Curver Performance Evaluation 
```{r }
auc <- rbind(performance(pr, measure = 'auc')@y.values[[1]],
             performance(pr2, measure = 'auc')@y.values[[1]],
             performance(pr3, measure = 'auc')@y.values[[1]])
rownames(auc) <- (c('Logistic Regression',  'CART','Random Forest'))
colnames(auc) <- 'Area Under ROC Curve'
round(auc, 4)

```

