---
title: "ML with R"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


#importing Libraries
```{r cars}
library(dplyr)
library(ggplot2)
library(caret)
library(caretEnsemble)
library(VIM)
#library(data.table)
library(DataExplorer) # For initial exploratory data analysis
```

#importing Dataset
```{r }
data<-read.csv(file.choose(), header=T)
```
# Overview of the dataset
```{r }
head(data)
```
Checking Diamension of the dataset
```{r }
dim(data)
```
#checking structure of dataset and their attribute data type
```{r }
str(data)
```
#Ploting structure of the dataset
```{r }
plot_str(data)
```
# Checking missing values in the dataset

```{r }
sapply(data, function(x) sum(is.na(x)))
```

#Ploting the missing values
```{r }
plot_missing(data)
```
#Proportion analysis of Martial Status attribute in the dataset
```{r }
prop.table(table(data$Marital_Status))
```
#Proportion analysis of Age attribute in the dataset

```{r }
prop.table(table(data$Age))
```
#Proportion analysis of City Category attribute in the dataset


```{r }
prop.table(table(data$City_Category))
```
##Proportion analysis of tay_In_Current_City_Years attribute in the dataset

```{r }
prop.table(table(data$Stay_In_Current_City_Years))

```
#ploting the dataset throught Bar chart
```{r }
plot_bar(data)
```
# Histogram of the dataset

```{r }
plot_histogram(data)
```
# Data type conversion of the dataset attributes
```{r }
data$User_ID <- as.factor(data$User_ID)
data$Product_ID <- as.factor(data$Product_ID)
data$Gender <- as.factor(if_else(data$Gender == 'M', 'Male', 'Female'))
data$Age <- as.factor(data$Age)
data$Occupation <- as.factor(data$Occupation)
data$City_Category <- as.factor(data$City_Category)
data$Stay_In_Current_City_Years <- as.factor(data$Stay_In_Current_City_Years)
data$Marital_Status <- as.factor(if_else(data$Marital_Status == 1, 'Married', 'Sin
gle'))
data$Product_Category_1 <- as.integer(data$Product_Category_1)
data$Product_Category_2 <- as.integer(data$Product_Category_2)
data$Product_Category_3 <- as.integer(data$Product_Category_3)
data$Purchase <- as.numeric(data$Purchase)
```
# Dealing with Missing Values
```{r }
library(rpart)
fit <- rpart(Product_Category_2 ~ User_ID + Product_ID + Age + Gender,
data = data[!is.na(data$Product_Category_2),],
method = "anova")
data$Product_Category_2[is.na(data$Product_Category_2)] <-
predict(fit, data[is.na(data$Product_Category_2),])
fit_1 <- rpart(Product_Category_3 ~ User_ID + Product_ID + Age + Gender,
data = data[!is.na(data$Product_Category_3),],
method = "anova")
data$Product_Category_3[is.na(data$Product_Category_3)] <-
predict(fit_1, data[is.na(data$Product_Category_3),])
plot_missing(data)
```
# How many unique User_IDs are there in the dataset?
```{r }
length(unique(data$User_ID))
```
# How many items did each customer purchased?
```{r }
Unique_UserID <- as.data.frame(table(data$User_ID))
names(Unique_UserID) <- c("User_ID", "Customer_Purchase_Count")
head(Unique_UserID)
```

```{r }
new_data <- data %>%
group_by(User_ID, Age, Gender, Occupation, City_Category, Stay_In_Current_City_Years, Marital_Status) %>%
summarise_each(funs(mean), Product_Category_1, Product_Category_2, Product_Category_3, Purchase)
```
# Rename the average values accordingly
```{r }
colnames(new_data)[8] <- "Product_Cat_1_Avg"
colnames(new_data)[9] <- "Product_Cat_2_Avg"
colnames(new_data)[10] <- "Product_Cat_3_Avg"
colnames(new_data)[11] <- "Avg_Purchase_Amount"
```
# Explore the age and gender variables versus the product categories and the purchase amount
# 1. Which Age group/gender had the highest purchase by product category?
# Product Category 1:
```{r }
ggplot(new_data, aes(Gender, Product_Cat_1_Avg, fill = Gender)) + geom_col(width = 0.4) + facet_wrap(~ Age) +
labs(title = "Age Group/Gender Vs Product Category 1")
```
# Product Category 2:
```{r }
ggplot(new_data, aes(Gender, Product_Cat_2_Avg, fill = Gender)) + geom_col() + facet_wrap(~ Age)+labs(title = "Age Group/Gender Vs Product Category 2")
```
# Product Category 3:
```{r }
ggplot(new_data, aes(Gender, Product_Cat_3_Avg, fill = Gender)) + geom_col() + facet_wrap(~ Age)+labs(title = "Age Group/Gender Vs Product Category 3")
```
# Age group versus Average purchase amount
```{r }
ggplot(new_data, aes(Age, Avg_Purchase_Amount, fill = Gender)) + geom_col() + facet_wrap(~ Gender) +labs(title = "Age Group/Gender Vs Average Purchase Amount")
```
#Histogram of Purchase Column by Marital Status
```{r }
ggplot(data, aes(x = Purchase, fill = Marital_Status)) +
  geom_histogram(bins = 75) +
  facet_wrap(~ Marital_Status) +
  labs(title= "Purchases Histogram by Marital Status") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
#Histogram of Purchase Column by Stay in Current City
```{r }
ggplot(data, aes(x = Purchase, fill = Stay_In_Current_City_Years)) +
  geom_histogram(bins = 75) +
  facet_wrap(~ Stay_In_Current_City_Years) +
  labs(title= "Purchases Histogram by Stay In Current City Years") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
#Histogram of Purchase Column by City Category
```{r }
ggplot(data, aes(x = Purchase, fill = City_Category)) +
  geom_histogram(bins = 75) +
  facet_wrap(~ City_Category) +
  labs(title= "Purchases Histogram by City Category") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
# Removing uncessary Attribute
```{r }
black_friday <- data %>%
  select(-User_ID, -Product_ID)

summary(black_friday)
```

```{r }
head(black_friday)
```
# Data Partition
```{r }
set.seed(123)
which_train <- createDataPartition(black_friday$Purchase, 
                                   p = 0.8, 
                                   list = FALSE) 
friday_train <- black_friday[which_train,]
friday_test <- black_friday[-which_train,]
```

# Linear Regression Machine Learning Model
```{r }
linear_train <-  train(Purchase ~ ., data = friday_train, method = "lm")
summary(linear_train)
```
#Predict the test set and compute RMSE

```{r }
linear_train_forecasts <- predict(linear_train, friday_test)
linear_RMSE <- RMSE(linear_train_forecasts, friday_test$Purchase)
linear_RMSE
```
# Compute Compute R-square
```{r }
linear_Rsq <-  R2(linear_train_forecasts,  friday_test$Purchase)
linear_Rsq
```
# Vector Machine Regression(SVM) and Setting train control

```{r }

ctrl_cv2 <- trainControl(method = "cv",
                         number = 2)

# Grid search
parametersC <- data.frame(C = c(0.01, 0.1, 0.2, 0.5, 1, 5))
```
# Train data and Summary
```{r }

set.seed(123)
svm_Linear <- train(Purchase ~ ., 
                    data = friday_train, 
                    method = "svmLinear",
                    tuneGrid = parametersC,
                    trControl = ctrl_cv2)


summary(svm_Linear)
```
#Predict the test set and compute RMSE and Compute the prediction error, RMSE
```{r }

svm_train_forecasts <- predict(svm_Linear, friday_test)
svm_RMSE <- RMSE(svm_train_forecasts, friday_test$Purchase)
svm_RMSE
```